---
title: "Missing Data Mechanisms and Multiple Imputation with miceFast"
author: "Maciej Nasinski"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Missing Data Mechanisms and Multiple Imputation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Introduction

Handling missing data is one of the most common challenges in applied statistics.
The validity of any imputation strategy depends critically on **why** the data are
missing. This vignette introduces the three missing-data mechanisms —
MCAR, MAR, and MNAR - and then demonstrates how to use **miceFast** for proper
Multiple Imputation (MI) with Rubin's rules.

For the full API reference (all imputation models, `fill_NA()`, `fill_NA_N()`, and
the OOP interface), see the companion vignette:
[miceFast Introduction and Advanced Usage](miceFast-intro.html).

```{r packages}
library(miceFast)
library(dplyr)
library(ggplot2)
set.seed(2025)
```

---

# Missing-Data Mechanisms

Little and Rubin (2002) defined three mechanisms that describe the relationship
between the data and the probability of missingness:

## MCAR — Missing Completely at Random

A variable is **MCAR** when the probability of being missing is the same for all
observations, regardless of both observed and unobserved values.

$$P(R = 0 \mid Y_{obs}, Y_{mis}) = P(R = 0)$$

where $R$ is the missingness indicator, $Y_{obs}$ are the observed values, and
$Y_{mis}$ are the missing values.

**Example:** A sensor fails at random intervals due to a hardware glitch unrelated
to the measurements.

**Testing for MCAR:**
MCAR is the only mechanism that can be tested from the observed data.
Little's (1988) test compares the means of observed values across
different missingness patterns; a significant result rejects MCAR.
However, failure to reject does not prove MCAR — the test has limited
power, especially with small samples or few patterns. It is a necessary
check, not a sufficient one.

**Implications:**

- Complete-case analysis (listwise deletion) produces unbiased estimates but is
  less efficient.
- Simple imputation methods (mean, median) are unbiased for the mean but distort
  the variance and covariance structure.
- MI is valid and efficient.

## MAR — Missing at Random

A variable is **MAR** when the probability of being missing depends on **observed**
data but not on the missing values themselves, conditional on the observed data.

$$P(R = 0 \mid Y_{obs}, Y_{mis}) = P(R = 0 \mid Y_{obs})$$

**Example:** Patients with higher blood pressure (recorded) are less likely to
return for a follow-up cholesterol measurement. Cholesterol is missing depending
on blood pressure, not on the cholesterol value itself.

**Implications:**

- Complete-case analysis is generally biased.
- Model-based imputation conditioning on observed predictors yields valid inferences.
- MI with an appropriate imputation model is the standard approach.
- **miceFast** models (`lm_bayes`, `lm_noise`, `lda`) condition on observed predictors,
  making them appropriate for MAR.

**MAR cannot be tested.**
Unlike MCAR, there is no test that can distinguish MAR from MNAR using the
observed data alone, because the distinction depends on the unobserved values.
The practical response is to make MAR more plausible by including **auxiliary
variables** — variables that predict missingness or the missing values even
if they are not part of the analysis model. The more information in the
imputation model, the weaker the assumption required.

Note also that the boundary between MAR and MNAR shifts depending on which
variables are conditioned on. A variable that appears MNAR with a sparse
predictor set may be MAR once the right covariates are included (see Collins
et al., 2001).

## MNAR — Missing Not at Random

A variable is **MNAR** when the probability of being missing depends on the
**unobserved** (missing) value itself, even after conditioning on observed data.

$$P(R = 0 \mid Y_{obs}, Y_{mis}) \neq P(R = 0 \mid Y_{obs})$$

**Example:** People with very high income are less likely to report their income
in a survey — the missingness depends on the value that would have been reported.

**Implications:**

- No imputation method can fully correct the bias without additional assumptions
  or external information.
- Sensitivity analysis is crucial: impute under MAR assumptions, then examine how
  results change under plausible MNAR deviations.
- Selection models may be needed.

## Practical guidance

| Mechanism | Testable? | Imputation valid? | Strategy |
|-----------|-----------|------------------|----------|
| MCAR | Yes (Little's test) | Yes | Any method works; MI is most efficient |
| MAR | No | Yes (if model is correct) | Condition on observed predictors; use MI |
| MNAR | No | Partially | Use MI as a starting point + sensitivity analysis |

In practice, you cannot distinguish MAR from MNAR using the observed data alone.
Including many predictors of missingness in the imputation model makes the MAR
assumption more plausible and reduces the sensitivity to violation.
Always perform sensitivity analyses.

---

# Multiple Imputation: Theory

Multiple Imputation (MI), introduced by Rubin (1987), addresses the fundamental
problem of single imputation: it fails to account for the **uncertainty** introduced
by not knowing the true missing values. For a comprehensive modern treatment
of MI, see van Buuren (2018).

## The MI procedure

1. **Impute** — Create $m$ completed datasets by drawing from the predictive
   distribution of the missing values (each draw uses a different random seed or
   stochastic model).
2. **Analyze** — Fit the analysis model of interest (e.g., `lm`, `glm`) separately
   on each of the $m$ completed datasets.
3. **Pool** — Combine the $m$ sets of estimates and standard errors using
   **Rubin's rules**.

## Rubin's rules

Let $\hat{Q}_j$ and $\hat{U}_j$ denote the point estimate and variance estimate
from the $j$-th completed dataset ($j = 1, \ldots, m$).

**Pooled estimate:**
$$\bar{Q} = \frac{1}{m} \sum_{j=1}^{m} \hat{Q}_j$$

**Within-imputation variance:**
$$\bar{U} = \frac{1}{m} \sum_{j=1}^{m} \hat{U}_j$$

**Between-imputation variance:**
$$B = \frac{1}{m-1} \sum_{j=1}^{m} (\hat{Q}_j - \bar{Q})^2$$

**Total variance:**
$$T = \bar{U} + \left(1 + \frac{1}{m}\right) B$$

The degrees of freedom for inference use the **Barnard-Rubin adjustment**
(Barnard & Rubin, 1999):

$$\nu = (m - 1)\left(1 + \frac{1}{r}\right)^{2}$$

where $r = (1 + 1/m)B / \bar{U}$ is the **relative increase in variance** due to
missingness. With finite complete-data degrees of freedom $\nu_{com}$, the adjusted
degrees of freedom are:

$$\nu_{adj} = \left(\frac{1}{\nu} + \frac{1}{\hat{\nu}_{obs}}\right)^{-1}$$

where $\hat{\nu}_{obs} = \nu_{com}(\nu_{com} + 1)(1 - \gamma) / (\nu_{com} + 3)$
and $\gamma = (1 + 1/m)B / T$ is the **proportion of total variance** due to
missingness (denoted $\lambda$ in the output of `pool()`).

### Diagnostic quantities

- **$\lambda$ (lambda):** Proportion of total variance due to missingness: $\lambda = (1 + 1/m)B / T$.
- **FMI:** Fraction of missing information, approximating the loss of information due to missing data.
- **RIV:** Relative increase in variance: $r = (1 + 1/m)B / \bar{U}$.

### Why MI works

Each imputed dataset is a draw from the posterior predictive distribution
$P(Y_{mis} \mid Y_{obs})$. The between-imputation variance $B$ captures
the additional uncertainty from not knowing the missing values. Rubin (1987)
showed that this procedure yields valid frequentist inference — confidence
intervals with correct coverage — provided two conditions hold: the
imputations are **proper** and the imputation model is **congenial** with the
analysis model. The efficiency of MI depends on the **fraction of missing
information** (FMI), not on the fraction of missing data. A dataset
with 40% missing values may have low FMI if the observed predictors are highly
informative.

### Proper imputation

An imputation procedure is **proper** (Rubin, 1987) if it propagates all
sources of uncertainty: both the uncertainty about the model parameters and
the residual variability. In practice this means each imputation must
(1) draw model parameters from their posterior distribution, and then
(2) draw imputed values from the predictive distribution given those parameters.

`lm_bayes` in miceFast is a proper imputation method: it draws
$\beta$ and $\sigma^2$ from the posterior under a non-informative prior,
then generates imputed values from the resulting predictive distribution.
`lm_noise`, which adds residual noise to point-estimate predictions, is
**improper** — it omits the parameter uncertainty and tends to produce
confidence intervals that are slightly too narrow, though the bias is small
when $n$ is large relative to $p$ (Schafer, 1997, Section 3.4).

### Congeniality

The imputation model is **congenial** with the analysis model (Meng, 1994)
when the imputation model is at least as general as — or richer than — the
analysis model. If the analysis model includes an interaction term but the
imputation model does not, the resulting MI inferences can be biased.
Practically: include in the imputation model all variables that will appear
in the analysis model, plus any auxiliary variables that predict missingness.
Err on the side of a richer imputation model.

---

# Multiple Imputation with miceFast

## Why miceFast for MI?

The **mice** package provides a complete MI framework but its imputation step can be slow,
especially with large datasets, many groups, or many imputations. **miceFast** provides
fast C++/Armadillo implementations of the same underlying models, which can be plugged
into the standard MI workflow:

| Step | mice | miceFast |
|------|------|----------|
| 1. Impute | `mice()` | `fill_NA()` in a loop |
| 2. Analyze | `with()` | `lapply()` + model fitting |
| 3. Pool | `mice::pool()` | `miceFast::pool()` |

The `pool()` function in **miceFast** implements Rubin's rules with the Barnard-Rubin
degrees-of-freedom adjustment. It has been tested against `mice::pool()` for
linear, logistic, and Poisson regression.

## Stochastic models for MI

For MI to work correctly, each imputation must introduce appropriate **random variation**.
Deterministic models (like `lm_pred`) should *not* be used for MI because they produce
identical imputations across datasets.

| Variable type | Recommended model | How it's stochastic | Proper? |
|--------------|-------------------|---------------------|---------|
| Continuous | `lm_bayes` | Draws regression coefficients from their posterior distribution | Yes |
| Continuous | `lm_noise` | Adds residual noise drawn from the estimated error distribution | No (omits parameter uncertainty) |
| Categorical | `lda` + random `ridge` | Different ridge penalties across imputations create variation | Approximate |

## Basic MI workflow

```{r mi-basic}
data(air_miss)

# Step 1: Create m = 10 completed datasets
m <- 10
completed <- lapply(1:m, function(i) {
  air_miss %>%
    mutate(
      Ozone_imp = fill_NA(
        x = ., model = "lm_bayes",
        posit_y = "Ozone",
        posit_x = c("Solar.R", "Wind", "Temp")
      )
    )
})

# Step 2: Fit the analysis model on each
fits <- lapply(completed, function(d) {
  lm(Ozone_imp ~ Wind + Temp, data = d)
})

# Step 3: Pool using Rubin's rules
pool_result <- pool(fits)
pool_result

# Detailed summary with confidence intervals
summary(pool_result)
```

The output includes:

- **estimate**: Pooled coefficient ($\bar{Q}$).
- **std.error**: Square root of total variance ($\sqrt{T}$).
- **statistic**: Wald test statistic.
- **df**: Barnard-Rubin adjusted degrees of freedom.
- **p.value**: Two-sided p-value.
- **lambda / fmi**: Fraction of total variance / missing information due to missingness.
- **conf.low / conf.high**: 95% confidence intervals.

## MI with mixed variable types

Real datasets often have both continuous and categorical variables with missing values.
Use `lm_bayes` for continuous variables and `lda` with a random `ridge` for categorical variables:

```{r mi-mixed}
data(air_miss)

impute_dataset <- function(data) {
  data %>%
    mutate(
      # Continuous: Bayesian linear regression
      Ozone_imp = fill_NA(
        x = ., model = "lm_bayes",
        posit_y = "Ozone",
        posit_x = c("Solar.R", "Wind", "Temp")
      ),
      Solar_R_imp = fill_NA(
        x = ., model = "lm_bayes",
        posit_y = "Solar.R",
        posit_x = c("Wind", "Temp", "Intercept")
      ),
      # Categorical: LDA with random ridge for stochasticity
      Ozone_chac_imp = fill_NA(
        x = ., model = "lda",
        posit_y = "Ozone_chac",
        posit_x = c("Wind", "Temp"),
        ridge = runif(1, 0.5, 50)
      )
    )
}

set.seed(777)
completed <- replicate(10, impute_dataset(air_miss), simplify = FALSE)

# Fit and pool a model for continuous outcome
fits_ozone <- lapply(completed, function(d) {
  lm(Ozone_imp ~ Wind + Temp + Solar_R_imp, data = d)
})
pool(fits_ozone)
```

## MI with GLMs

`pool()` works with any model that supports `coef()` and `vcov()`. Here's an example
with logistic regression:

```{r mi-glm}
data(air_miss)

# Create a binary outcome for demonstration
completed <- lapply(1:10, function(i) {
  d <- air_miss %>%
    mutate(
      Ozone_imp = fill_NA(
        x = ., model = "lm_bayes",
        posit_y = "Ozone",
        posit_x = c("Solar.R", "Wind", "Temp")
      )
    )
  d$high_ozone <- as.integer(d$Ozone_imp > median(d$Ozone_imp, na.rm = TRUE))
  d
})

fits_logit <- lapply(completed, function(d) {
  glm(high_ozone ~ Wind + Temp, data = d, family = binomial)
})

pool(fits_logit)
```

## MI with grouped imputation

When the relationship between variables differs across groups, fit separate
imputation models per group:

```{r mi-grouped}
data(air_miss)

completed_grouped <- lapply(1:5, function(i) {
  air_miss %>%
    group_by(groups) %>%
    do(mutate(.,
      Ozone_imp = fill_NA(
        x = ., model = "lm_bayes",
        posit_y = "Ozone",
        posit_x = c("Wind", "Temp", "Intercept")
      )
    )) %>%
    ungroup()
})

fits <- lapply(completed_grouped, function(d) {
  lm(Ozone_imp ~ Wind + Temp + factor(groups), data = d)
})
pool(fits)
```

## MI with weighted imputation

If your data have sampling weights or you want to account for heteroscedasticity:

```{r mi-weighted}
data(air_miss)

completed_w <- lapply(1:5, function(i) {
  air_miss %>%
    mutate(
      Solar_R_imp = fill_NA(
        x = ., model = "lm_bayes",
        posit_y = "Solar.R",
        posit_x = c("Wind", "Temp", "Intercept"),
        w = weights
      )
    )
})

fits_w <- lapply(completed_w, function(d) {
  lm(Solar_R_imp ~ Wind + Temp, data = d)
})
pool(fits_w)
```

---

# Sensitivity Analysis

## Why sensitivity analysis?

Since we can never be certain whether data are MAR or MNAR, it is important to check
whether substantive conclusions are robust to different imputation approaches.

## Comparing models with `fill_NA_N()`

`fill_NA_N()` averages over *k* imputed values, making it useful for quick sensitivity
checks. Compare different imputation strategies on a single dataset:

```{r sensitivity-models}
data(air_miss)

air_sensitivity <- air_miss %>%
  mutate(
    Ozone_bayes = fill_NA(x = ., model = "lm_bayes",
      posit_y = "Ozone", posit_x = c("Solar.R", "Wind", "Temp")),
    Ozone_noise = fill_NA(x = ., model = "lm_noise",
      posit_y = "Ozone", posit_x = c("Solar.R", "Wind", "Temp")),
    Ozone_pmm = fill_NA_N(x = ., model = "pmm",
      posit_y = "Ozone", posit_x = c("Solar.R", "Wind", "Temp"), k = 20),
    Ozone_pred = fill_NA(x = ., model = "lm_pred",
      posit_y = "Ozone", posit_x = c("Solar.R", "Wind", "Temp"))
  )

compare_imp(air_sensitivity, origin = "Ozone",
  target = c("Ozone_bayes", "Ozone_noise", "Ozone_pmm", "Ozone_pred"))
```

## Varying the number of imputations

Check whether results stabilize as *m* increases:

```{r sensitivity-m}
results <- data.frame()

for (m_val in c(3, 5, 10, 20, 50)) {
  completed <- lapply(1:m_val, function(i) {
    air_miss %>%
      mutate(Ozone_imp = fill_NA(
        x = ., model = "lm_bayes",
        posit_y = "Ozone", posit_x = c("Solar.R", "Wind", "Temp")
      ))
  })
  fits <- lapply(completed, function(d) lm(Ozone_imp ~ Wind + Temp, data = d))
  s <- summary(pool(fits))
  s$m <- m_val
  results <- rbind(results, s)
}

results %>%
  filter(term == "Wind") %>%
  ggplot(aes(x = m, y = estimate)) +
  geom_point(size = 3) +
  geom_errorbar(aes(ymin = conf.low, ymax = conf.high), width = 2) +
  labs(title = "Stability of Wind coefficient across m",
       x = "Number of imputations (m)",
       y = "Pooled estimate") +
  theme_minimal()
```

## Comparing with naive imputation

Demonstrate how naive (mean) imputation underestimates uncertainty compared to
proper MI:

```{r sensitivity-naive}
data(air_miss)

# Naive: fill_NA with lm_pred + Intercept only = mean imputation
air_naive <- air_miss %>%
  mutate(
    Ozone_mean = fill_NA(
      x = ., model = "lm_pred",
      posit_y = "Ozone",
      posit_x = "Intercept"
    )
  )

fit_naive <- lm(Ozone_mean ~ Wind + Temp, data = air_naive)
ci_naive <- confint(fit_naive)

# Proper MI
completed <- lapply(1:20, function(i) {
  air_miss %>%
    mutate(Ozone_imp = fill_NA(
      x = ., model = "lm_bayes",
      posit_y = "Ozone", posit_x = c("Solar.R", "Wind", "Temp")
    ))
})
fits <- lapply(completed, function(d) lm(Ozone_imp ~ Wind + Temp, data = d))
pool_summary <- summary(pool(fits))

cat("Naive imputation CI for Wind:\n")
print(ci_naive["Wind", ])

cat("\nMI pooled CI for Wind:\n")
cat(sprintf("  2.5%%: %.4f  97.5%%: %.4f\n",
    pool_summary$conf.low[pool_summary$term == "Wind"],
    pool_summary$conf.high[pool_summary$term == "Wind"]))
```

---

# Choosing the Number of Imputations

Rubin (1987) showed that with $m$ imputations the efficiency relative to
$m = \infty$ is approximately $(1 + \lambda/m)^{-1}$, where $\lambda$ is the
FMI. The original recommendation of $m = 5$ assumed FMI below 50% and
was aimed at point estimates. For hypothesis tests and confidence intervals,
more imputations are needed: White et al. (2011) suggested $m \geq 20$, and
von Hippel (2020) proposed a two-stage rule: run a pilot with small $m$,
estimate FMI, then set $m$ so that the Monte Carlo error of the MI estimate
is small relative to its standard error.

Since **miceFast** imputation is fast, there is little reason to be stingy.
Use $m = 20$ or more, and increase $m$ until estimates and standard errors
stabilize (see the sensitivity analysis above).

---

# Practical Checklist

Before imputing, examine the missingness pattern (`upset_NA()`) and check
collinearity (`VIF()` — values above 10 are problematic). Include auxiliary
variables that predict missingness even if they are not in the analysis model.

For continuous variables use `lm_bayes` or `lm_noise`; for categorical
variables use `lda` with a random `ridge`; for right-skewed data consider
`logreg = TRUE`. Use at least $m = 20$ imputations, pool with `pool()`,
and run sensitivity analyses varying the model and $m$.

Report the imputation model, number of imputations, and any assumptions
about the missing-data mechanism.

---

# Comparison with mice

| Feature | mice | miceFast |
|---------|------|----------|
| MI framework | Complete (FCS/MICE algorithm) | Building blocks for MI |
| Imputation models | 25+ built-in methods | `lm_pred`, `lm_bayes`, `lm_noise`, `lda`, `pmm` |
| Chained equations | Yes (iterative multivariate) | Manual chaining possible |
| Speed | R-based | C++/Armadillo — significantly faster |
| Grouping | Via `blocks` | Built-in, auto-sorted |
| Weights | Limited | Full support |
| Pooling | `mice::pool()` | `miceFast::pool()` |
| Diagnostics | Trace plots, convergence | `compare_imp()`, `upset_NA()` |

**mice** (van Buuren & Groothuis-Oudshoorn, 2011) is the right choice when you
need the full chained-equations algorithm, passive imputation, or its diagnostic
tooling.
**miceFast** is useful when speed matters (large data, many groups, many imputations)
or when you want imputation inside `dplyr`/`data.table` pipelines.
The two can be combined: use **miceFast** for imputation and **mice** for diagnostics, or vice versa.

---

# References

- Rubin, D.B. (1987). *Multiple Imputation for Nonresponse in Surveys*. John Wiley & Sons.

- Barnard, J. and Rubin, D.B. (1999). Small-sample degrees of freedom with
  multiple imputation. *Biometrika*, 86(4), 948-955.

- Collins, L.M., Schafer, J.L., and Kam, C.-M. (2001). A comparison of inclusive
  and restrictive strategies in modern missing data procedures. *Psychological
  Methods*, 6(4), 330-351.

- Little, R.J.A. (1988). A test of missing completely at random for multivariate
  data with missing values. *Journal of the American Statistical Association*,
  83(404), 1198-1202.

- Little, R.J.A. and Rubin, D.B. (2002). *Statistical Analysis with Missing Data*
  (2nd ed.). John Wiley & Sons.

- Meng, X.-L. (1994). Multiple-imputation inferences with uncongenial sources of
  input. *Statistical Science*, 9(4), 538-558.

- Schafer, J.L. (1997). *Analysis of Incomplete Multivariate Data*. Chapman & Hall/CRC.

- van Buuren, S. (2018). *Flexible Imputation of Missing Data* (2nd ed.). Chapman & Hall/CRC.
  [Online version](https://stefvanbuuren.name/fimd/).

- van Buuren, S. and Groothuis-Oudshoorn, K. (2011). mice: Multivariate Imputation
  by Chained Equations in R. *Journal of Statistical Software*, 45(3), 1-67.

- White, I.R., Royston, P., and Wood, A.M. (2011). Multiple imputation using chained
  equations: Issues and guidance for practice. *Statistics in Medicine*, 30(4), 377-399.

- von Hippel, P.T. (2020). How many imputations do you need? A two-stage calculation
  using a quadratic rule. *Sociological Methods & Research*, 49(3), 699-718.
